Optimal Decomposition-Recombination Plans for 2D Well-Constrained Graphs

### 2014-12-05 doc/GMP2015/s01_introduction.tex
s01_introduction.tex edited online with Bitbucket

\section{Introduction}
Modeling, analysis and design of quasi-uniform or self-similar layered materials using geometric constraint systems

Some properties of natural and engineered materials can be studied by treating them as stacked two dimensional (2D) layers, 
where the structure within each layer may be recursively 
self-similar \cite{Intro1}, 
spanning multiple scales. Within any one scale, the material layer is generally aperiodic, and quasi-uniform,
consisting of a few repeated motifs appearing in disordered arrangments.
Each layer is not necessarily planar, i.e., it consists of multiple, inter-constraining planar (genus 0) monolayers. 
Furthermore, a layer is often   either isostatic or somewhat flexible, and rarely self-stressed, at any scale. These properties are
consistent with a self-assembled 2D structure that minimizes mass, optimally distributes external stresses and itself 
participates in the assembly of diverse and multifunctional, 
larger 2D structures.

@@

In order to study structural and mechanical properties such as rigidity, flexes and distribution of external stresses 
within the material layer, it is natural to model a material layer as a solution of geometric constraint systems 
arising from appropriate geometric primitives under metric or algebraic constraints. 
Such  2D {\em qusecs} (quasi-uniform or self-similar constraint systems) can be used to 
to understand or design material layers (their solutions) with desired properties.

@@

Fast, graph-based algorithms exist (pebble-game \cite{XX}),   
for locating all maximal, generically rigid subsystems \cite{XX}. 
When the input is rigid, these algorithms do nothing, i.e, the output is the same as the input.
However, both for self-similar or just aperiodic 2D qusecs, it is imperative to recursively decompose rigid systems 
into their rigid subsystems, down to the level of geometric primitives, in order to 
understand or design properties at all scales, such as: rigidity, flexes, distribution 
of external stresses as well as robustness/sensitivity to constraint variations.

Recursive decomposition of geometric constraint systems has been formalized \cite{XX} and well-studied \cite{XX} 
as the Decomposition-Recombination (DR)-planning problem.
For the above-mentioned classes of 2D qusecs, generic rigidity is 
a combinatorial property and hence decomposition can, in principle, be done by a graph-based algorithm, without
involving geometric information in the constraint system.  
However, since many  such decompositions can exist for a given constraint system, 
criteria defining desirable or optimal DR-plans and DR-planning algorithms were given in \cite{XX}. An {\em optimal DR-plan} is one
that minimizes the maximum number of child subsystems of any parent system, a property that is 
crucial for obtaining a solution to the parent constraint system from the solutions of the child systems.

For a special class of 2D qusecs, namely  {\em tree-decomposable} systems \cite{XX} common in computer aided mechanical design, 
(which includes ruler-and-compass and Henneberg-I constuctible systems), all DR-plans turn out to 
be optimal. This satisfies the so-called {\em Church-Rosser} criterion, leading to highly efficient DR-planning algorithms.
However, for general 2D qusecs, even when restricted to bar-joint systems, 
the optimal DR-planning problem was shown to be NP-hard \cite{XX}.
Subsequently, alternate criteria were suggested   
such as {\em cluster minimality} 
requiring parent systems to be composed of  
a minimal set of at least 2 rigid proper subsystems (i.e., no proper subset forms a rigid system); 
and {\em proper maximality}, requiring child subsystems
to be maximal rigid proper subsystems of the parent system.
While polynomial time algorithms were given \cite{XX} to generate DR-plans meeting the cluster minimality criterion,
no such algorithm is known for the latter criterion.

For fairly general constraint systems, there are algorithms that use a constraint system's DR-plan \cite{XX} for solving the system.
They optimize the algebraic complexity of (re)combining child system solutions, 
and algebraically solve the optimized recombination system to give a solution to the parent system at any
given level of the DR-plan. However, the generality of these algorithms trades-off with efficiency. 
Faster algorithms should exist for 2D cusecs that leverage recent developments on so-called Cayley configuration spaces
\cite{XX,XX}.

For underconstrained 2D cusecs obtained from various subclasses of tree-decomposable systems, algorithms have been developed \cite{XX} that 
leverage Cayley configuration space theory to provide descriptions of connected components and find paths within the connected components 
of the standard cartesian configuration spaces. However, no such algorithms are known for non-tree-decomposable systems.

\subsection{Contributions}
In this paper, we  navigate the NP-hardness barrier for finding optimal DR-plans by defining a 
so-called {\it canonical} DR-plan and showing that all canonical DR-plans for isostatic 2DGMCS are optimal. 
We then give an efficient ($O(n^2)$) algorithm to find a canonical DR-plan for all 2DGMCS. The notion of a canonical DR-plan 
illuminates the essence of the NP-hardness of finding optimal DR-plans for overconstrained 2DGMCS and gives classes of 
overconstrained 2DGMCS for which the above  algorithm finds an optimal DR-plan. We then show how to judiciously combine 
existing methods for recombination, search of so-called, convex Cayley configuration spaces, and navigation of realizations, 
to explore a connected component of the configuration space of a flexible 2DGMCS, given a starting configuration. 
Software implementation can be found at: \todo{Provide Link}.

>>>>

\section{Introduction}
\label{sec:prev}
Modeling, analysis and design of quasi-uniform or self-similar layered materials using geometric constraint systems

Some properties of natural and engineered materials can be studied by treating them as stacked two dimensional (2D) layers, 
where the structure within each layer may be recursively 
self-similar \cite{Intro1}, 
spanning multiple scales. Within any one scale, the material layer is generally aperiodic, and quasi-uniform,
consisting of a few repeated motifs appearing in disordered arrangments.
Each layer is not necessarily planar, i.e., it consists of multiple, inter-constraining planar (genus 0) monolayers. 
Furthermore, a layer is often  either {\sl isostatic or flexible, not self-stressed}, at any scale. These properties are
consistent with a self-assembled 2D structure that minimizes mass, optimally distributes external stresses and itself 
participates in the assembly of diverse and multifunctional, 
larger 2D structures.

@@

In order to study structural and mechanical properties such as rigidity, flexes and distribution of external stresses 
within the material layer, it is natural to model a material layer as a solution of geometric constraint systems 
arising from appropriate geometric primitives under metric or algebraic constraints. 
Such  2D {\bf{\em qusecs}} (quasi-uniform or self-similar constraint systems) can be used to 
to understand or design material layers (their solutions) with desired properties.

@@

\medskip\noindent\underbar{{\sl (1) Finding (vertex)-maximal generically rigid subsystems}}
Fast, graph-based algorithms exist (pebble-game \cite{XX}),   
for locating all maximal, generically rigid subsystems \cite{XX}. 
When the input is rigid, these algorithms do nothing, i.e, the output is the same as the input.
However, both for self-similar or just aperiodic 2D qusecs, it is imperative to recursively decompose rigid systems 
into their rigid subsystems, down to the level of geometric primitives, in order to 
understand or design properties at all scales, such as: rigidity, flexes, distribution 
of external stresses as well as robustness/sensitivity to constraint variations.

\medskip\noindent\underbar{{\sl (2) Optimal Recursive Decomposition (DR-planning)}}
Recursive decomposition of geometric constraint systems has been formalized \cite{XX} and well-studied \cite{XX} 
as the {\sl Decomposition-Recombination (DR)-planning} problem.
For the above-mentioned classes of 2D qusecs, generic rigidity is 
a combinatorial property and hence each level of the decomposition should, in principle, be achievable by a graph-based algorithm as in (1), without
involving geometric information in the constraint system.  
Since many  such decompositions can exist for a given constraint system, 
criteria defining desirable or optimal DR-plans and DR-planning algorithms were given in \cite{XX}. An {\em optimal DR-plan} is one
that minimizes the maximum number of child subsystems of any parent system, a property that is 
crucial for obtaining a solution to the parent constraint system from the solutions of the child systems.
However, for general 2D qusecs, even when restricted to bar-joint systems, 
the optimal DR-planning problem was shown to be NP-hard \cite{XX}.

\medskip\noindent\underbar{{\sl (3) DR-plans for special classes and with other criteria}}
For a special class of 2D qusecs, namely  {\em tree-decomposable} systems \cite{XX} common in computer aided mechanical design, 
(which includes ruler-and-compass and Henneberg-I constuctible systems), all DR-plans turn out to 
be optimal. This satisfies the so-called {\em Church-Rosser} criterion, leading to highly efficient DR-planning algorithms.
Subsequently, alternate criteria were suggested   
such as {\em cluster minimality} 
requiring parent systems to be composed of  
a minimal set of at least 2 rigid proper subsystems (i.e., no proper subset forms a rigid system); 
and {\em proper maximality}, requiring child subsystems
to be maximal rigid proper subsystems of the parent system.
While polynomial time algorithms were given \cite{XX} to generate DR-plans meeting the cluster minimality criterion,
no such algorithm is known for the latter criterion.

\medskip\noindent\underbar{{\sl (4) Optimal Recombination and Solution Space Navigation}}
For fairly general 3D constraint systems, there are algorithms with formal guarantees that 
solve the parent (sub)systems that occur in the DR-plan \cite{XX, XX,XX}, 
provided the DR-plan meets some of the above criteria.
These algorithms  uncover underlying matroids to combinatorially optimize the algebraic complexity of (re)combining child system solutions, 
and algebraically solve the optimized recombination system to give a solution to the parent system at any
given level of the DR-plan. However, the generality of these algorithms trades-off against efficiency, and despite the optimization,
the best algorithms can still take
exponential time in the number of child subsystems in order to guarantee all solutions even for a single (sub)system in a DR-plan, 
and are prohibitively slow in practice. For the entire DR-plan, finding all solutions is barely tractable even with fast recombination algorithms
for the subsystems in the DR-plan. This is because even for the simplest systems, the problem is NP-hard \cite{XX} and there is a 
combinatorial explosion of solutions \cite{XX}.  Using the DR-plan and optimal recombination as a principled basis, high performance software exists \cite{XX,XX} to 
tame combinatorial explosion via user intervention.

\medskip\noindent\underbar{{\sl (5) Configuration Spaces of Underconstrained Systems}}
For underconstrained 2D bar-joint and multibody-pin cusecs obtained from various subclasses of tree-decomposable systems, algorithms have been developed \cite{XX, XX} that 
to complete them into isostatic systems \cite{XX, XX} and to find paths within the connected components \cite{XX,XX} of
standard cartesian configuration spaces. Most of the algorithms with formal guarantees leverage Cayley configuration space theory \cite{XX,XX}
to characterize subclasses of graphs and additional constraints that control combinatorial explosion, and provide faithful bijective representation 
of connected components and paths.
These algorithms have decreasing efficiency as the subclass of systems gets bigger, starting with
underlying partial 2-tree graphs (alternately, tree-width 2, series-parallel, $K_4$ minor avoiding),  moving on   
to 1 degree-of-freedom (dof) graphs with low Cayley complexity (which include common linkages such as the Strandbeest, Limacon and Cardioid), 
and further to general 1-dof tree-decomposable graphs. 
While software suites exist \cite{XX,XX}, no such formal algorithms and guarantees are known for non-tree-decomposable systems.

\subsection{Contributions}
\label{sec:cont}
In this paper, we  navigate the NP-hardness barrier mentioned in (2) above, for finding optimal DR-plans by defining a 
so-called {\it canonical} DR-plan and showing that all canonical DR-plans for isostatic 2D qusecs are optimal. 
We then give an efficient ($O(n^2)$) algorithm to find a canonical DR-plan for all 2DGMCS. The notion of a canonical DR-plan 
illuminates the essence of the NP-hardness of finding optimal DR-plans for overconstrained 2DGMCS and gives classes of 
overconstrained 2DGMCS for which the above  algorithm finds an optimal DR-plan. We then show how to judiciously combine 
existing methods for recombination, search of so-called, convex Cayley configuration spaces, and navigation of realizations, 
to explore a connected component of the configuration space of a flexible 2DGMCS, given a starting configuration. 
Software implementation can be found at: \todo{Provide Link}.
